Other metrics

### 2021-02-25 analysis/rev_prototyping/additional_metrics_results.Rmd

mcdb_new_di <-read.csv(here::here("mcdb_new_di.csv"))

all_di <- all_di %>%
  filter(s0 > 1,
         s0 != n0,
         n0 != s0 + 1,
         !singletons)

@@

@@

@@

>>>>

all_di <- all_di %>%
  mutate(log_nparts = log(gmp:::as.double.bigz(nparts)),
         log_nsamples = log(nsamples),
         log_s0 = log(s0),
         log_n0 = log(n0)) %>%
  filter(n0 != s0,
         s0 != 1,
         !singletons,
         n0 != (s0 + 1)) %>%
  mutate(dat = ifelse(grepl(dat, pattern = "fia"), "fia", dat),
         dat = ifelse(dat == "misc_abund_short", "misc_abund", dat)) %>%
  mutate(Dataset = dat,
         Dataset = ifelse(Dataset == "fia", "FIA", Dataset),
         Dataset = ifelse(Dataset == "bbs", "Breeding Bird Survey", Dataset),
         Dataset = ifelse(Dataset == "mcdb", "Mammal Communities", Dataset),
         Dataset = ifelse(Dataset == "gentry", "Gentry", Dataset),
         Dataset = ifelse(Dataset == "misc_abund", "Misc. Abundance", Dataset)) %>%
  filter(nparts > 20)

@@

All of these currently exclude FIA because those are still running.

@@

Visualization and analysis of nsingletons has a little more nuance than the others, because there are often relatively few values for nsingletons at all. 80% of sites have fewer than 20 singletons as the 95th percentile, which is just a rough way of saying that a **lot** of these are going to be sensitive to whether you define the percentiles as > or >=.

In general the strict > percentile will give you an (appropriately) conservative estimate of how many are extraordinarily **high** and the >= will give you an appropriate estimate of how many are unusually **low**. For most metrics it doesn't really matter, writ large, which you use, because ties are rare. For this one, you get large numbers of sites where a lot of values are = to the observed values, and the >= decision will therefore give you a jump of a lot of percentile scores.

For the "proportions high/low" calculations, we use *the > percentile for high* and *the >= percentile for low*. For visualization, because we are interested in both unusually high and unusually low scores, we can't just pick one or the other. The histogram using >= is reliable at the high end but has a misleading spike at 0, and vice versa. I am making these plots using the *mean*, which doesn't have the misleading spikes at the extremes but does smear things out a little bit.

```{r, fig.dim = c(5,5)}

nsingletons_results <- all_di %>%
  group_by_all() %>%
  mutate(singletons_percentile_jump = nsingletons_percentile - nsingletons_percentile_excl,
         nsingletons_percentile_mean = sum(nsingletons_percentile, nsingletons_percentile_excl) / 2) %>%
  ungroup() %>%
  mutate(prop_singles_actual = nsingletons / s0,
         sim_prop_singles_95 = nsingletons_95 / s0,
         sim_prop_singles_mean = nsingletons_mean / s0) %>%
  mutate(actual_minus_95_prop = prop_singles_actual - sim_prop_singles_95) %>%
  mutate(high_singles=nsingletons_percentile > 95,
         high_singles_ex = nsingletons_percentile_excl > 95,
         low_singles = nsingletons_percentile < 5,
         low_singles_ex = nsingletons_percentile_excl < 5) %>%
  mutate(singles_desc = ifelse(high_singles_ex, "high", ifelse(low_singles, "low", "int")))

BBS, MCDB, and Misc have too many rare species, while Gentry has too few:

```{r}
ggplot(filter(nsingletons_results, nparts > 20), aes(nsingletons_percentile_mean)) +
  geom_histogram(bins = 100) +
  facet_wrap(vars(dat), scales = "free_y", ncol = 1) +
  geom_vline(xintercept = 95)

nsingletons_results  %>% 
  group_by(dat) %>%
  filter(nparts > 20) %>%
  summarize(            prop_nsingletons_high_raw = mean(nsingletons_percentile_excl > 95),
         #   prop_nsingletons_mean_high_raw = mean(nsingletons_percentile_mean_actual > 95),
      #      prop_nsingletons_mean_high_jk = mean(nsingletons_percentile_mean > 95),
            prop_nsingletons_low_raw = mean(nsingletons_percentile  < 5),
            nsites_included = dplyr::n())

_Of sites where the nsingletons value is extreme_, we're looking at - for example, in MCDB - an increase of .2 in the proportion of singleton species in the observed vector over the 95th percentile of scores for the proportion of singleton species for the feasible set:

nsingletons_results <- nsingletons_results %>%
  mutate(singletons_change = (nsingletons - nsingletons_95) / s0)

ggplot(filter(nsingletons_results, high_singles_ex), aes(singletons_change)) +
  geom_histogram() +
  facet_wrap(vars(dat), scales = "free")

nsingletons_results %>%
  filter(high_singles_ex) %>%
  group_by(dat) %>%
  summarize(mean_singles_change = mean(singletons_change),
            n = dplyr::n())

Broken out by N/S, you can see that the low values are concentrated in the Gentry where N/S < 3, and that high values are most common as N/S > 10:

ggplot(nsingletons_results, aes(n0/s0, nsingletons_percentile_excl, color = singles_desc)) +
  geom_point(data = filter(nsingletons_results, singles_desc != "int")) +
  geom_point(alpha = .1, color = "grey") +
  facet_wrap(vars(dat), scales = "free") +
  scale_x_log10() +
  geom_vline(xintercept = c(3, 10))

For N/S < 3:

```{r}
nsingletons_results %>%
  filter(n0/s0 < 3) %>%
  group_by(dat, singletons) %>%
  summarize(
            prop_high_singles_ex = mean(high_singles_ex),
            prop_low_singles = mean(low_singles),
            nsites= dplyr::n()) %>%
  ungroup()

For N/S  > 10:

nsingletons_results %>%
  filter(n0/s0 > 10) %>%
  group_by(dat, singletons) %>%
  summarize(
            prop_high_singles_ex = mean(high_singles_ex),
            prop_low_singles = mean(low_singles),
            nsites= dplyr::n()) %>%
  ungroup()

Working on a breadth index here (stuff needs to run).

@@

Most datasets have very low Shannon diversity, except for Gentry:

ggplot(filter(all_di, nparts > 20), aes(shannon_percentile)) +
  geom_histogram(bins = 100) +
  facet_wrap(vars(dat), scales = "free_y", ncol = 1) +
  geom_vline(xintercept = 5)

all_di  %>% 
  group_by(dat) %>%
  filter(nparts > 20) %>%
  summarize(
            prop_shannon_high_raw = mean(shannon_percentile > 95),
            prop_shannon_low_raw = mean(shannon_percentile < 5),
            nsites_included = dplyr::n())
```

Working on the breadth index (stuff needs to run).
